---
title: "STA302 Part 1"
author: "Yixiao Zhang"
output: pdf_document
---
# The packages and the dataset 
We are going to use datasets from the World Health Organization data repository.

```{r, eval=TRUE, echo = T}
library(rms)
library(knitr)
library(tidyverse)
library(broom) ## Needed to make the regression output 'tidy'
library(ggplot2)
library(xtable)
library(glmnet)
library(car)
library(Matrix)
library(MPV)
library(MASS)
library(gglasso)
library(pls)
library(psych)
library(ggpubr)
```

```{r}
LED <- read.csv("Life_Expectancy_Data.csv", header = T)
```

Let's clean the data a bit:
```{r}
temp_data <- LED %>% dplyr::select(Country, Year, Life.expectancy, infant.deaths, Alcohol, Total.expenditure, BMI, HIV.AIDS)
cleaned_data <- na.omit(temp_data)
colnames(cleaned_data)[3] <- "LE"

```
Before we begin let's split the data 70/30 into training and testing:

```{r}
set.seed(1006787700)

## use 70% as training 30% as test
sample <- sample(c(TRUE, FALSE), nrow(cleaned_data), replace=TRUE, prob=c(0.7,0.3))
train  <- cleaned_data[sample, ]
test   <- cleaned_data[!sample, ]


```

```{r}
library(ggplot2)
# Let's begin by calculating and saving correlation matrix
corr_matrix <- cor(cleaned_data[,4:8])

# Melt correlation matrix to long format
melted_correlation <- reshape2::melt(corr_matrix)

melted_correlation

# Next let's plot the correlation
ggplot(data = melted_correlation) + 
  geom_tile(aes(x = Var1, y = Var2, fill = value)) +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = 0, name = "Correlation") +
  labs(title = "Correlation heatmap", subtitle = "Variables from cleaned dataset") +
  xlab("Variable 1") + ylab("Variable 2") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        axis.text.y = element_text(angle = 0, vjust = 0.5, hjust=1),
        panel.background = element_rect(fill = "white"),
        panel.grid.major = element_line(colour = "gray90"),
        panel.grid.minor = element_blank(),
        legend.title = element_blank(),
        legend.key.width = unit(1, "cm"),
        legend.key.height = unit(0.5, "cm"),
        legend.text = element_text(size = 10),
        plot.title = element_text(size = 18, face = "bold"))
```

Residual plot:
```{r}
model <- lm(LE ~ infant.deaths + Alcohol + Total.expenditure + BMI + HIV.AIDS, data = train)
hii <- hatvalues(model)
which(hii > 4/nrow(train))
## The standardized residuals ##
r <- rstudent(model)
which(r <= -2 | r >= 2)
r[which(hii > 4/nrow(train))]

### Plot using Base R ###
par(family = 'serif')
residual_plot <- plot(fitted(model), r, type = "p", xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Standardized Residuals", col = "blue")
abline(h = 2, lty = 2)
abline(h = -2, lty = 2)

ggarrange(residual_plot, ncol = 2, nrow = 2)



```

QQplot:
```{r}
### Fit the model and get the levarage points ##

plot(fitted(model), r, type = "p", xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Standardized Residuals", col = ifelse(r>=2 | r<=-2, "red", "blue"))
abline(h = 2, lty = 2)
abline(h = -2, lty = 2)
text(fitted(model)[r>=2 | r<=-2]+0.5, r[r>=2 | r<=-2], labels = which(r <= -2 | r >= 2))

### now that we observe the leverage points, we can clean the data by removing them: ##
outliers <- which(r <= -2 | r >= 2)
data_no_outliers <- cleaned_data[-outliers,]
model_no_outliers <- lm(LE ~ infant.deaths + Alcohol + Total.expenditure + BMI + HIV.AIDS, data = data_no_outliers)
fitted_no_outliers <- fitted(model_no_outliers)
### model without outlier


# qq plot
r1 <- rstudent(model_no_outliers)
p1 <- ggplot(data_no_outliers, aes(sample = r1)) + theme_minimal() 
p1 + stat_qq() + stat_qq_line()

# F - test:
summary(model)

# transformation
boxcox(model, lambda = seq(-4, 4, 1/10) )
boxcox(lm(infant.deaths + 1 ~ 1, data = train))
boxcox(lm(Alcohol ~ 1, data = train))
boxcox(lm(Total.expenditure ~ 1, data = train))
boxcox(lm(BMI ~ 1, data = train))
boxcox(lm(HIV.AIDS ~ 1, data = train))

```
So optimal lambda for life.expectancy, infant.deaths, Alcohol, Total.expenditure, BMI, HIV.AIDS are 2, -0.5, 0.5, 0.5, 1,  -0.5

```{r}
# Performing coxbox we have:
life.expectancy_new <- (train$LE^2 - 1) / 2

infant.deaths_new <- log(train$infant.deaths + 1)

Alcohol_new <- (train$Alcohol^(0.5) - 1) * 2
Total.expenditure_new <- (train$Total.expenditure^(0.5) - 1) * 2
BMI_new <- (train$BMI - 1)
HIV.AIDS_new <- ((train$HIV.AIDS)^(-0.5) - 1) * -2

transformed_train <- data.frame(life.expectancy_new, infant.deaths_new, Alcohol_new, Total.expenditure_new, BMI_new, HIV.AIDS_new)

model_transformed <- lm(life.expectancy_new ~ infant.deaths_new + Alcohol_new + Total.expenditure_new + BMI_new + HIV.AIDS_new, data = transformed_train)


# residual plot 
hii <- hatvalues(model_transformed)
which(hii > 4/nrow(train))
## The standardized residuals ##
r <- rstudent(model_transformed)
which(r <= -2 | r >= 2)
r[which(hii > 4/nrow(train))]

### Plot using Base R ###
par(family = 'serif')
plot(fitted(model_transformed), r, type = "p", xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Standardized Residuals", col = "blue")
abline(h = 2, lty = 2)
abline(h = -2, lty = 2)

plot(fitted(model_transformed), r, type = "p", xlab = "Fitted Values", ylab = "Standardized Residuals",
     main = "Standardized Residuals", col = ifelse(r>=2 | r<=-2, "red", "blue"))
abline(h = 2, lty = 2)
abline(h = -2, lty = 2)
text(fitted(model)[r>=2 | r<=-2]+0.5, r[r>=2 | r<=-2], labels = which(r <= -2 | r >= 2))

## remove outliers
outliers <- which(r <= -2 | r >= 2)
data_no_outliers <- train[-outliers,]

## tranformation
life.expectancy_new <- (data_no_outliers$LE^2 - 1) / 2

infant.deaths_new <- (data_no_outliers$infant.deaths^(-0.5) - 1) * -2

Alcohol_new <- (data_no_outliers$Alcohol^(0.5) - 1) * 2
Total.expenditure_new <- (data_no_outliers$Total.expenditure^(0.5) - 1) * 2
BMI_new <- (data_no_outliers$BMI - 1)
HIV.AIDS_new <- (data_no_outliers$HIV.AIDS^(-0.5) - 1) * -2

## fit the model to data without outliers:
model_no_outliers <- lm(life.expectancy_new ~ infant.deaths + Alcohol_new + Total.expenditure_new + BMI_new + HIV.AIDS_new, data = data_no_outliers)
fitted_no_outliers <- fitted(model_no_outliers)


# qq plot
r1 <- rstudent(model_no_outliers)
p1 <- ggplot(data_no_outliers, aes(sample = r1)) + theme_minimal() 
p1 + stat_qq() + stat_qq_line()



```



Given the 5 variables I choose in the proposal，we will make sure that none are greater than 5. 
```{r}
# VIF for all predictors 
summary(model_transformed)
v <- vif(model_transformed)



# None of the variable has vif greater than 5 so this is good

```
Variable selection: aic, bic, lasso:

```{r, eval=TRUE, echo = T}
####### Variable selection #######
# Performing boxcox we have:
life.expectancy_new <- (train$LE^2 - 1) / 2


infant.deaths_new <- (log(train$infant.deaths + 1) )

Alcohol_new <- (train$Alcohol^(0.5) - 1) * 2
Total.expenditure_new <- (train$Total.expenditure^(0.5) - 1) * 2
BMI_new <- (train$BMI - 1)
HIV.AIDS_new <- (train$HIV.AIDS^(-0.5) - 1) * -2
variable <- data.frame(life.expectancy_new, infant.deaths_new, Alcohol_new, Total.expenditure_new, BMI_new, HIV.AIDS_new)
## Step wise regression

## Based on AIC ##
model.lm <- lm(life.expectancy_new ~ Alcohol_new + Total.expenditure_new + BMI_new + HIV.AIDS_new, data = variable)
summary(model.lm)  
n <- nrow(train)
sel.var.aic <- step(model.lm, trace = 0, k = 2, direction = "both") 
sel.var.aic<-attr(terms(sel.var.aic), "term.labels")   
sel.var.aic

## Based on BIC ##
model.lm <- lm(life.expectancy_new ~  Alcohol_new + Total.expenditure_new + BMI_new + HIV.AIDS_new, data = variable)
summary(model.lm)   
n <- nrow(variable)
sel.var.bic <- step(model.lm, trace = 0, k = log(n), direction = "both") 
sel.var.bic<-attr(terms(sel.var.bic), "term.labels")   
sel.var.bic

### LASSO selection 

## Perform cross validation and choose lambda 
set.seed(1006787700)
cv.out <- cv.glmnet(x = as.matrix(variable), y = train$LE, standardize = T, alpha = 1)
plot(cv.out)
best.lambda <- cv.out$lambda.1se
best.lambda
co<-coef(cv.out, s = "lambda.1se")

#Selection of the significant features(predictors)

## threshold for variable selection

thresh <- 0.00
# select variables 
inds<-which(abs(co) > thresh )
variables<-row.names(co)[inds]
sel.var.lasso<-variables[!(variables %in% '(Intercept)')]
sel.var.lasso
```


```{r, eval=TRUE, echo = T}
library(rms)
set.seed(1006787700)

### Based on AIC 
ols.aic <- ols(life.expectancy_new ~ ., data = variable[,which(colnames(variable) %in% c(sel.var.aic, "life.expectancy_new"))], 
               x=T, y=T, model = T)

## 10 fold cross validation    
aic.cross <- calibrate(ols.aic, method = "crossvalidation", B = 10)

## Calibration plot
plot(aic.cross, las = 1, xlab = "Predicted Life Expectancy", main = "Cross-Validation calibration with AIC")




### Based on BIC 
ols.bic <- ols(life.expectancy_new ~ ., variable[,which(colnames(variable) %in% c(sel.var.bic, "life.expectancy_new"))], 
               x=T, y=T, model = T)

## 10 fold cross validation    
bic.cross <- calibrate(ols.bic, method = "crossvalidation", B = 10)
## Calibration plot
plot(bic.cross, las = 1, xlab = "Predicted Life Expectancy", main = "Cross-Validation calibration with BIC")

### Based on Lasso 
ols.lasso <- ols(life.expectancy_new ~ ., data = variable[,which(colnames(variable) %in% c(sel.var.bic, "life.expectancy_new"))], 
                 x=T, y=T, model = T)

## Cross validation 
lasso.cross <- calibrate(ols.lasso, method = "crossvalidation", B = 10)
## Calibration plot
plot(lasso.cross, las = 1, xlab = "Predicted Life Expectancy", main = "Cross-Validation calibration with LASSO")


## Calculate Train Error 
pred.aic <- predict(ols.aic, newdata = variable[,which(colnames(variable) %in% c(sel.var.aic, "life.expectancy_new"))])
## Prediction error 
AIC_prediction_error <- mean((variable$life.expectancy_new - pred.aic)^2)
AIC_prediction_error


## Calculate Train Error 
pred.bic <- predict(ols.bic, newdata = variable[,which(colnames(variable) %in% c(sel.var.bic, "life.expectancy_new"))])
## Prediction error ##
BIC_prediction_error <- mean((variable$life.expectancy_new - pred.bic)^2)
BIC_prediction_error


## Calculate Train Error 
pred.lasso <- predict(ols.lasso, newdata = variable[,which(colnames(variable) %in% c(sel.var.lasso, "life.expectancy_new"))])
## Prediction error 
lasso_prediction_error <- mean((variable$life.expectancy_new - pred.lasso)^2)
lasso_prediction_error

```
lasso has the lowest abs mean error, however, it might be overly simplistic with just two predictors, so I will choose bic

```{r}

## Check assumptions


model <- lm(life.expectancy_new ~ ., variable[,which(colnames(variable) %in% c(sel.var.bic, "life.expectancy_new"))]) 
## Calculate h_ii ##
hii <- hatvalues(model)

## Identify the leverages & bad leverage points ##
r <- rstudent(model)
bad_leverage_point <-which(r <= -2 | r >= 2)


## Standardized residuals ##
r <- rstudent(model)

## Checking assumptions
plot(fitted(model), r, type = "p", xlab = "Life Expectancy After Transformation", ylab = "Standardized Residuals",
     main = "Standardized Residuals", col = ifelse(r>=2 | r<=-2, "red", "blue"),ylim=range(-5,5))
abline(h = 2, lty =2)
abline(h = -2, lty = 2)

qqnorm(rstudent(model),col = "blue")
qqline(rstudent(model))

# Vif values:
vif_values <- vif(model)
vif_values


pairs(variable[,which(colnames(variable) %in% c(sel.var.bic, "life.expectancy_new"))],col = "blue", main = "Pairs plot")

```
Now we remove outliers and refit the model
```{r}
data_no_outliers <- variable[-bad_leverage_point,]
final_model <- lm(life.expectancy_new ~ ., variable[,which(colnames(variable) %in% c(sel.var.bic, "life.expectancy_new"))]) 

#output the model:
summary(final_model)

## Test Error ##

##First transform the test data:
life.expectancy_new <- (test$LE^2 - 1) / 2


infant.deaths_new <- (log(test$infant.deaths + 1) )

Alcohol_new <- (test$Alcohol^(0.5) - 1) * 2
Total.expenditure_new <- (test$Total.expenditure^(0.5) - 1) * 2
BMI_new <- (test$BMI - 1)
HIV.AIDS_new <- (test$HIV.AIDS^(-0.5) - 1) * -2

test_transformed <- data.frame(life.expectancy_new, infant.deaths_new, Alcohol_new, Total.expenditure_new, BMI_new, HIV.AIDS_new)

## make prediction

pred <- predict(final_model, newdata = test_transformed[,which(colnames(test_transformed) %in% c(sel.var.bic, "life.expectancy_new"))])
## Prediction error 
test_error <- mean((test_transformed$life.expectancy_new - pred)^2)


table <- matrix(c(BIC_prediction_error, test_error), nrow = 1)
colnames(table)<- list("BIC - Training Pred error", "BIC - Test Pred error")
rownames(table)<- "Prediction Error"

kable(table, caption = "Training vs Test prediction error"
        )
```







